---
title: "Weak instruments"
author: "Keith Barnatchez"
date: "May 6, 2020"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(mvtnorm)
```

As a first exercise, let's look at how the "strength" of our instrument and the relationship between our variable of interest and the confounding factor impact our results when using regular vs. two-stage least squares. You can use the sliders below to alter both parameters. The subsequent figure is the result of drawing samples of increasing size under the assumptions you impose with the sliders.

```{r, echo=FALSE}

inputPanel(
  
  sliderInput("corr_x1_x2", label = "Confounding correlation:",
              min = -0.95, max = 0.95, value = 0.05, step = 0.01),
  
  sliderInput("corr_x1_z", label = "Instrument correlation:",
              min = -0.95, max = 0.95, value = 0.9, step = 0.01)
)

renderPlot({
  minN <- 50
  maxN <- 1000
  
  cor_x1_x2 <- input$corr_x1_x2
  cor_x1_z <- input$corr_x1_z
  cor_x2_z <- 0.0
  
  sigma <- matrix(c(1,cor_x1_x2,cor_x1_z,
                    cor_x1_x2,1,cor_x2_z,
                    cor_x1_z,cor_x2_z,1),ncol = 3)
  
  beta_2s <- rep(NA,maxN-minN+1) 
  beta    <- rep(NA,maxN-minN+1)

    # Loop 
  i <- 1
  for (n in minN:maxN) {
  
    X <- rmvnorm(n,mean=c(0,0,0),sigma=sigma, method="eigen")
    x1 <- X[,1]
    x2 <- X[,2]
    z <-  X[,3]
    
    y <- 0.3*x1 - 0.7*x2 + rnorm(n) 
    xhat <- lm(x1 ~ z)$fitted.values
    
    beta[i] <- lm(y ~ x1)$coefficients[2]
    beta_2s[i] <- lm(y ~ xhat)$coefficients[2]
    i <- i+1
  
  }
  
  # Make dfs
  df_raw <- data.frame(n=minN:maxN,beta=beta,Version=rep('Unadjusted',maxN-minN+1))
  df_2s  <-  data.frame(n=minN:maxN,beta=beta_2s,Version=rep('Two-stage',maxN-minN+1))
  df <- rbind(df_raw,df_2s)
  
  
  # Plot
  
  ggplot(df, aes(x=n,y=beta,color=Version)) + geom_line() + geom_hline(yintercept = 0.3)  +
    scale_y_continuous(limits=c(-0.5,1))

})

```

