---
title: Riddler January 15
author: Keith Barnatchez
date: '2021-01-16'
slug: riddler-january-15
categories:
  - Riddler
tags: []
subtitle: ''
summary: ''
authors: []
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---
```{r, echo=FALSE}
library(ggplot2)
```

> You’re a contestant on the hit new game show, “You Bet Your Fife.” On the show, a random real number (i.e., decimals are allowed) is chosen between 0 and 100. Your job is to guess a value that is less than this randomly chosen number. Your reward for winning is a novelty fife that is valued precisely at your guess. For example, if the number is 75 and you guess 5, you’d win a 5 dollar fife, but if you’d guessed 60, you’d win a 60 dollar fife. Meanwhile, a guess of 80 would win you nothing.
>
> What number should you guess to maximize the average value of your fifing winnings?

## Simulation 

First, we'll approximate the solution via simulation.

```{r simulation}
nsim <- 1e5
rewards <- rep(NaN,100)

for (n in 1:100) { # loop over (discrete) guesses
  cumreward <- 0
  for (s in 1:nsim) { # loop through simulated rewards
    if (n < 100*runif(1)) {
      cumreward <- cumreward+n
    }
  } # for s 
  rewards[n] <- cumreward/nsim 
} # for n
```
After storing the results, we can plot the average reward over time:
```{r plot, echo=FALSE}
tempdf <- data.frame(rewards, 1:100)
colnames(tempdf) <- c('reward','guess')

ggplot(tempdf, aes(x=guess,y=reward)) + geom_line(color='darkblue', size=2) +
         labs(x='Guess',
            y='Avg reward',
            title='Average rewards over different guesses',
            subtitle='Rewards peak at 50') + theme_bw() 
```

## Analytic approach

Let's represent our rewards from playing the game by the random variable $r$, and $n$ denote our choice of number between 0 and 100. We care about the conditional distribution of rewards for a given number, $p(r|n)$. This is a discrete distribution, as there are only two possibilities: win $n$ with probability $1-\frac{n}{100}$, and nothing with probability $\frac{n}{100}$. In turn, we can write out the conditional expectation of $r$:
$$
\mathbb{E}(r|n) = n\left(1-\frac{n}{100}\right) + 0 \times \frac{n}{100} =  n-\frac{n^2}{100}
$$
We want to choose $n$ to maximize this expression:
$$
\begin{aligned}
\frac{\partial \mathbb{E}(r|n)}{\partial n}=0 & \iff 1-\frac{n}{50}=0 \\
& \iff n=50
\end{aligned}
$$